---
title: "Data analysis assignment 2"
author: "Nina Cunningham"
date: "11th February 2020"
output: github_document
---

```{r setup, include=FALSE}
# Please note these options.
# This tells R Markdown that we want to show code in the output document.
knitr::opts_chunk$set(echo = TRUE)
# Switching off messages in the output document. 
knitr::opts_chunk$set(message = FALSE)
# Switching on caching to make things faster (don't commit cache files on Github).
knitr::opts_chunk$set(cache = TRUE)
```

In this assignment you will work with relational data, i.e. data coming from different data tables that you can combine using keys. Please read ch.13 from R for Data Science before completing this assignment -- https://r4ds.had.co.nz/relational-data.html.

## Read data

We will work with three different tables: household roster from wave 8 (*h_egoalt*), stable characteristics of individuals (*xwavedat*), and household data from wave 8 (*h_hhresp*).

```{r}
library(tidyverse)
# You need to complete the paths to these files on your computer.
Egoalt8 <- read_tsv("C:/Users/ninaj/Documents/datan3/datan3-master/UKDA-6614-tab/tab/ukhls_w8/h_egoalt.tab")
Stable <- read_tsv("C:/Users/ninaj/Documents/datan3/datan3-master/UKDA-6614-tab/tab/ukhls_wx/xwavedat.tab")
Hh8 <- read_tsv("C:/Users/ninaj/Documents/datan3/datan3-master/UKDA-6614-tab/tab/ukhls_w8/h_hhresp.tab")
```

## Filter household roster data (10 points)

The **egoalt8** data table contains data on the kin and other relationships between people in the same household. In each row in this table you will have a pair of individuals in the same household: ego (identified by *pidp*) and alter (identified by *apidp*). *h_relationship_dv* shows the type of relationship between ego and alter. You can check the codes in the Understanding Society codebooks here -- https://www.understandingsociety.ac.uk/documentation/mainstage/dataset-documentation.

First we want to select only pairs of individuals who are husbands and wives or cohabiting partners (codes 1 and 2). For convenience, we also want to keep only the variables *pidp*, *apidp*, *h_hidp* (household identifier), *h_relationship_dv*, *h_esex* (ego's sex), and *h_asex* (alter's sex).

```{r}
Partners8 <- Egoalt8 %>%
        filter(h_relationship_dv == 1 | h_relationship_dv == 2) %>%
        select(pidp, apidp, h_hidp, h_relationship_dv, h_sex, h_asex)
```

Each couple now appears in the data twice: 1) with one partner as ego and the other as alter, 2) the other way round. Now we will only focus on heterosexual couples, and keep one observation per couple with women as egos and men as their alters.

```{r}
Hetero8 <- Partners8 %>%
        # filter out same-sex couples and keep only one observation per couple with women as egos
        filter(h_sex == 2 & h_asex == 1)
```

## Recode data on ethnicity (10 points)

In this assignment we will explore ethnic endogamy, i.e. marriages and partnerships within the same ethnic group. First, let us a create a version of the table with stable individual characteristics with two variables only: *pidp* and *racel_dv* (ethnicity).

```{r}
Stable2 <- Stable %>%
        select(pidp, racel_dv)
```

Let's code missing values on ethnicity (-9) as NA.

```{r}
Stable2 <- Stable2 %>%
        mutate(racel_dv = recode(racel_dv, `-9` = NA_real_))
```

Now let us recode the variable on ethnicity into a new binary variable with the following values: "White" (codes 1 to 4) and "non-White" (all other codes).

```{r}
Stable2 <- Stable2 %>%
        mutate(race = recode(racel_dv,
                `1` = "White",
                `2` = "White",
                `3` = "White",
                `4` = "White",
                .default = "non-White"))
```

## Join data (30 points)

Now we want to join data from the household roster (*Hetero8*) and the data table with ethnicity (*Stable2*). First let us merge in the data on ego's ethnicity. We want to keep all the observations we have in *Hetero8*, but we don't want to add any other individuals from *Stable2*.

```{r}
JoinedEthn <- Hetero8 %>%
  left_join(Stable2, by = "pidp")
       
```

Let us rename the variables for ethnicity to clearly indicate that they refer to egos.

```{r}
JoinedEthn <- JoinedEthn %>%
        rename(egoRacel_dv = racel_dv) %>%
        rename(egoRace = race)
```

Now let us merge in the data on alter's ethnicity. Note that in this case the key variables have different names in two data tables; please refer to the documentation for your join function (or the relevant section from R for Data Science) to check the solution for this problem.

```{r}

JoinedEthn <- JoinedEthn %>%
  left_join(Stable2, by = c("apidp" = "pidp"))

```

Renaming the variables for alters.

```{r}
JoinedEthn <- JoinedEthn %>%
        rename(alterRacel_dv = racel_dv) %>%
        rename(alterRace = race)
```

## Explore probabilities of racial endogamy (20 points)

Let us start by looking at the joint distribution of race (White vs. non-White) of both partners.

```{r}
TableRace <- JoinedEthn %>%
        # filter out observations with missing data
        filter(!is.na(egoRace), !is.na(alterRace)) %>%
        count(egoRace, alterRace)
TableRace
```

Now calculate the following probabilities: 1) for a White woman to have a White partner, 2) for a White woman to have a non-White partner, 3) for a non-White woman to have a White partner, 4) for a non-White woman to have a non-White partner.

Of course, you can simply calculate these numbers manually. However, the code will not be reproducible: if the data change the code will need to be changed, too. Your task is to write reproducible code producing a table with the required four probabilities.


```{r}
TableRace %>%
        # group by ego's race to calculate sums
        group_by(egoRace) %>%
        # create a new variable with the total number of women by race
        mutate(total = sum(n)) %>%
        # create a new variable with the required probabilities 
        mutate(probability = n / total)
```

## Join with household data and calculate mean and median number of children by ethnic group (30 points)

1) Join the individual-level file with the household-level data from wave 8 (specifically, we want the variable for the number of children in the household).
2) Select only couples that are ethnically endogamous (i.e. partners come from the same ethnic group) for the following groups: White British, Indian, and Pakistani.
3) Produce a table showing the mean and median number of children in these households by ethnic group (make sure the table has meaningful labels for ethnic groups, not just numerical codes).
4) Write a short interpretation of your results. What could affect your findings?

```{r}
# recode Hh8 to make it easier
Hhk8 <- Hh8 %>%
  select(h_hidp, h_nkids_dv)

# Join individual-level file to household-level data
JoinedEthn2 <- JoinedEthn  %>%
  left_join(Hhk8, by = "h_hidp")

# Filter only ethically endogamous couples
JoinedEthn2 <- JoinedEthn2 %>%
  filter(egoRacel_dv == 1 & alterRacel_dv == 1 | egoRacel_dv == 9 & alterRacel_dv == 9 | egoRacel_dv == 10 & alterRacel_dv == 10)

# Recode to form a race catagory for easier understanding
JoinedEthn2 <- JoinedEthn2 %>%
  mutate(race = recode(egoRacel_dv,
                `1` = "White British",
                `9` = "Indian",
                `10` = "Pakistani"))

# Make a table for the number of children and edit the table to find mean and median
TableKids <- JoinedEthn2 %>%
  group_by(race) %>%
  summarise(mean = mean(h_nkids_dv, na.rm = TRUE), median = median(h_nkids_dv, na.rm = TRUE))

TableKids

``` 

The results show that the average number of children for White British is 0.5651314 (therefore 1 if you round up) with the median is 0 suggesting that the number of children per household is fairly low for White British parents. For Indian parents the mean is 0.9553753 (therefore 1 if you round up) and the median is 1. This suggest the number of children for Indian parents is larger than White British parents. However Pakistani households have a larger number of children with a mean of 1.8108747 (therefore 2 if you round up) and the median is 2. This suggests the Pakistani families have the largest number of children when compared to White British families (with the least) and Indian families.
However, many factors could affect this outcome. We removed missing values which could have skewed the result in a certain direction. There also could be a problem with the sample as certain people are selected, with the potential for non-response bias, and people may have dropped out of the study later on or misreport the number of children they have. Same-sex parents are also excluded from this table which could impact the results. Therefore, the results may not represent the actual number of children each race.

